\chapter{Transforming RISC-V to BTOR2}\label{chap:riscv_to_btor2}

\todo{Explain naming conventions for the model nodes}

This chapter addresses the central problem of this thesis:
transforming a RISC-V state into the BTOR2 format for benchmarking
purposes. F. Schrögendorfer conducted similar work in his master's
thesis \enquote{Bounded Model Checking in Lockless Programs}
\cite{bmcOfLockless}, where he describes, among other topics, an
encoding concept for a minimal machine in a multiprocessor context
\cite[Chapter 2]{bmcOfLockless}. In \cite[Chapter 8]{bmcOfLockless},
he outlines a method to encode programs for his machine model into a
BTOR2 model. This approach cannot be directly replicated here, as his
model assumes the entire program is known at encoding time, whereas I
aim to preserve the RISC-V property that allows for self-modifying
programs during execution. If this property were to be disregarded,
it would be possible to analyze the complete behavior of a program by
parsing the memory, but this is beyond the scope of this work. Even
so, I let myself inspire by his structuring of the model.

\section{The Concept}
To successfully execute a RISC-V instruction, three fundamental steps
must occur in sequence:
\begin{itemize}
    \item Fetch the current instruction from memory
    \item Identify the instruction
    \item Execute the instruction
\end{itemize}
Due to the fixed instruction length of RISC-V, as mentioned in \secref{sec:riscvIsa}, fetching the current instruction is straightforward. Ultimately, a node is required that retrieves a $word$ from memory at the location specified by $pc$.

For basic identification, the $opcode$ must be extracted and checked.
Depending on the opcode, further distinctions between instructions
require extracting and checking $funct3$ and, if necessary, $funct7$.
Ultimately, a node for each instruction is needed, holding a boolean
value indicating whether this instruction was fetched.

To execute the instruction, the values of the immediate $imm$ and, if
used, the registers $rs1$ and $rs2$ must be extracted. All
instructions only modify $rd$, $pc$, or memory. Therefore, the
next-state logic can be generalized for these three cases.

Memory is only modified when a store instruction is identified. As
all store instructions share the same type, computing the memory
address is consistent across them. The final step is overwriting the
memory at this address.

For the $pc$, except for jump commands, it always increments to point
to the next instruction. The two unconditional jumps, \texttt{JAL}
and \texttt{JALR}, must be handled separately. For branch
instructions, after determining whether the relevant condition for
the instruction holds, a general approach can be applied, as all
branch instructions execute the same operation from this point
onward.

With $rd$, generalization across instructions is not feasible.
However, it is possible to generalize across all possible registers
by adding a check in each register's update function to determine
whether the register in question is $rd$.

\section{Encoding}
For improved visualization in the BTOR2 code, all sort-IDs are marked
in \textcolor{UniGrey}{gray}, all node-IDs in
\textcolor{UniRed}{red}, and all non-ID numbers in
\textcolor{UniBlue}{blue}. As described in the BTOR2 syntax
\cite[Figure 1]{btor2}, each line can have an accompanying symbol.
Unfortunately, these cannot be used as aliases for the line numbers,
but for clarity, in the following figures I use them as such aliases.
This allows each new figure to start with the relative line number
\texttt{n}, making it feasible to describe processes with algorithms.
It is implied that \texttt{n} is sufficiently incremented after
adding to the model so that IDs do not overlap. The following
sections describe how a BTOR2 model is constructed from a RISC-V
state file.

\subsection{Constants}
First, I added the sorts and non-progressive constants needed in the
BTOR2 model, as shown in \figref{fig:constants}. This is extended by
a set of progressive constants used for comparison, e.g., against the
register number. \algoref{alg:progressiveconsts} describes how these
are added.

Of note is the representation of memory as an array of addressable
memory cells, each 1 byte. The chosen address space of 16 bits is
significantly smaller than the expected 64-bit address space, but
representing a 64-bit addressable memory with $2^{64}$ bytes
($\approx$ 18 exabytes) is not feasible. Therefore, I selected a
16-bit address space as a practical minimum, providing approximately
65kB and supporting programs with potentially over 10,000
instructions, which I consider sufficient for most use cases. The
encoding is implemented so that the address space can be modified as
needed. \todo{Change code to make address space modifiable by an
    option?} \todo{Explain progressive constants}

\input{figures/4-Transform/constants.tex}
\input{figures/4-Transform/progressiveconsts.tex}

\subsection{State Representation}
The next logical step is defining a representation of a RISC-V state.
This is straightforward, as shown in \figref{fig:states}. I also
introduced a flag for each register in my code to track whether the
register was written to, enabling the transformation of a witness to
a state file containing only the relevant registers. As these flags
do not affect the operation of the BTOR2 model and are only included
for an aesthetic choice, they are not included in my description and
will not be discussed further.
\input{figures/4-Transform/staterep.tex}

\subsection{Initialization}\label{sec:initialization}
To initialize a state in BTOR2 from a RISC-V state file, the values
in the registers must be loaded as constants, and for each memory
address mentioned in the state file, the value and address must be
loaded as constants. Due to the inability to represent a full 64-bit
address space, I must manage the reduction of the address space from
the state file to the BTOR2 model. I chose to initialize only the
addresses up to the BTOR2 model's address space maximum and omit all
others from the state file, as this provides the most predictable
behavior. All addresses not mentioned in the state file are
zero-initialized. Finally, these constants are used to initialize the
state. For the registers, this is straightforward; for memory, all
memory addresses are first written into a placeholder array, which is
then used to initialize the actual memory. Due to BTOR2 constraints,
these constants must be defined \textbf{before} the states, but
initialization with the values must occur after the states. Thus,
this initialization process \textbf{wraps around} the state
representation. The generation of constants is shown in
\algoref{alg:generateconstantsfromstate}, while the actual
initialization is shown in \algoref{alg:initstate}.
\input{figures/4-Transform/loadstateconstants.tex}
\input{figures/4-Transform/initstate.tex}

\subsection{Fetching the Current Instruction}
To fetch the current instruction, I read the four bytes of the
instruction and concatenate them, as shown in \figref{fig:fetching}.
\input{figures/4-Transform/fetching.tex}

\subsection{Deconstruction of the Instruction}
With the instruction available, it can be deconstructed to extract
the $opcode$, $rd$, $rs1$, $rs2$, $funct3$, $funct7$, and $imm$. For
everything except $imm$, this can be accomplished by shifting and
masking, as shown in \figref{fig:extractNOimm}.

The immediate, however, must first be constructed from its subfields,
which are referenced in \figref{fig:rv64i_formats}. In the BTOR2
model, this is shown in \figref{fig:extractimmbytype}.
\todo{Reference to same method in riscvsim} Three points are
noteworthy:\\ First, some immediate subfields overlap exactly. This
is utilized in lines (n + 1) with the overlap of $imm[11:5]$ for I-
and S-type, and (n + 21) with J- and B-types $imm[10:5]$ overlap.
Second, as described in \secref{sec:riscvIsa}, the immediate is
always sign-extended. This is achieved using arithmetic right shifts,
which perform sign extension and correctly position the highest
immediate bit. Third, at line (n + 8), sign extension requires a
right shift by 19. As this matches the opcode for arithmetic
instructions with immediate, I reused this constant.

Now, \textsl{iTypeImm}, \textsl{sTypeImm}, \textsl{bTypeImm},
\textsl{uTypeImm}, and \textsl{jTypeImm} are available. However, it
is preferable to have a single node \textsl{imm} referencing the
immediate value regardless of instruction. This is accomplished in
\figref{fig:findingImm}, where booleans are defined to check all
opcodes that are neither R-type nor I-type. Then, if-then-else nodes
are chained to select instructions of J-type, U-type, B-type, or
S-type. If the instruction is none of these, I default to I-type, as
R-type does not use an immediate value. Finally, $imm$ is extended to
the 64-bit width required by RV64I.

At this stage, the values of the designated $rs1$ and $rs2$ registers
can also be extracted. This is shown for $rs1$ in
\figref{alg:extractrs1val}; the process is identical for $rs2$, with
only the names changed. The starting equality comparisons can be
omitted for $rs2$, as they are already defined for $rs1$ and can be
referenced.

\input{figures/4-Transform/extractNOimm.tex}
\input{figures/4-Transform/extractimmbytype.tex}
\input{figures/4-Transform/choosecorrectimm.tex}
\input{figures/4-Transform/extractrs1val.tex}

\subsection{Instruction Detection}
For the next-state logic, it is essential to determine the current
command. Therefore, I defined a check
\textsl{is\textcolor{Green}{Instruction}} for each instruction. As
this is repetitive, \algoref{alg:commanddetection} describes a
generalized approach to obtain these booleans. An example for each
instruction subgroup in \algoref{alg:commanddetection} is provided in
\figref{fig:detectionexample}. The funct7 checks from the $needsf7$
subgroup can be reused if multiple instructions share the same
funct7.

\input{figures/4-Transform/commanddetection.tex}
\input{figures/4-Transform/detectionexample.tex}

\subsection{Next-State Logic}
The next-state logic is the core of the model. Almost everything else
supports this point. The goal is to create the changes each
instruction would make and then apply only the changes specific to
the instruction in the state. Each state node in the model must have
an accompanying next node to function correctly. First, the changed
values are computed.

\subsubsection{Creating All Values of Instruction Execution}
It is unnecessary to detail all instructions, as this simply follows
the RV64I ISA. Instead, I provide examples for each group of
instructions as divided in \tabref{tab:rv64i-instructions}. Examples
for \texttt{AUIPC}, \texttt{JALR}, \texttt{BEQ}, \texttt{LHU},
\texttt{SD}, \texttt{ANDI}, \texttt{SLLIW}, \texttt{SLT}, and
\texttt{SUBW} are shown in \figref{fig:valueexample}. These examples
illustrate overlaps that can be utilized, such as addresses for load
and store instructions or the 32-bit versions of word instructions.
The \texttt{SD} example demonstrates that all other store
instructions are interim results of preparing \texttt{SD}. Load
instructions are similar, but each requires sign extension to 64
bits.

With this, each change can be assigned to its instruction.

\input{figures/4-Transform/valueexample.tex}

\subsubsection{The Next Memory}
Defining the next memory array is straightforward. All store
instructions are cascaded through if-then-else nodes, with the final
'else' set as the current memory array; if no 'if' matches, the array
remains unchanged. This is shown in \figref{fig:nextmemory}.

\input{figures/4-Transform/nextmemory.tex}

\subsubsection{The Next pc}
For the next pc, the approach is similar, as shown in
\figref{fig:nextpc}. The only difference is that if no 'if' matches,
pc must point to the next instruction to execute. The nextPc value
was already computed for the JAL and JALR instructions and is reused
here. The unconditional jumps also modify the value in rd, which is
handled in the next section.

\input{figures/4-Transform/nextpc.tex}

\subsubsection{The Next rd}
At last, the remaining registers must be updated. The procedure is
defined in \figref{alg:nextrd}. With the exception of x0, this is the
same for all registers. The process is similar to defining the next
memory or pc, but instead of a handful of instructions, all 39
relevant instructions must be considered, as only branch and store
instructions do not modify rd. For brevity, the cascade for all
relevant instructions is not shown in full in \algoref{alg:nextrd},
but only indicated.

\input{figures/4-Transform/nextrd.tex}

\subsection{Constraints}
The final step is to define constraints to terminate the model
checker. The primary constraint is reaching a set number of
iterations, as shown in \figref{fig:badcounter}.

Additional constraints are defined to check for invalid instructions.
The first checks if the opcode is valid for the model. The second
constraint detects if the instruction cannot be identified even when
the opcode is valid, as shown in \figref{fig:unknownInstr}. The
constraint in \figref{fig:badaddress} handles
instruction-address-misaligned exceptions for jump instructions.

Other constraints can be defined, such as terminating on a specific
pc value or when a register reaches a specified value.

\todo{Maybe add examples on how to do this}\\ \todo{Maybe internal
    references in figures should be numbers\dots}

\input{figures/4-Transform/badcounter.tex}
\input{figures/4-Transform/badinstructions.tex}
\input{figures/4-Transform/badaddress.tex}

\section{Testing for Correctness}\label{sec:corectness}
To test my model, I compared its results to those of my RISC-V
simulator (\secref{sec:simulation}).

Given a state, both the simulation and the BTOR2 model are run with
the iteration maximum set to 1. The resulting BTOR2 witness cannot be
directly compared to the resulting state of the simulation.
Therefore, I implemented a simple converter from witness to state
\cite[src/restate\_witness.c]{repoRV2BTOR}. These two states can then
be compared. A shell script for this purpose is provided at
\cite[sh\_utils/compare\_iterations.sh]{repoRV2BTOR}.

To generate RISC-V states, I implemented a fuzzer
\cite[src/state\_fuzzer.c]{repoRV2BTOR} that generates randomized
states with one valid instruction at the address of pc. The fuzzer
first selects an instruction to test and fills all variable parts of
the instruction, such as $rd$ or $imm$. All registers relevant to the
instruction are then assigned random 64-bit values. A pc value is
generated to ensure the instruction fits within the limited address
space of the BTOR2 model. If a jump instruction is chosen, possible
address misalignment is corrected and address overflow is prevented.
This simplifies later comparison of the resulting states, as correct
execution of the instruction always results in the same state,
despite differences between the simulation and the BTOR2 model.

With this setup, a series of tests can be conducted. For this, I
implemented a shell script
\cite[sh\_utils/test\_btor2\_model.sh]{repoRV2BTOR}. As the number of
tests increases, it becomes more challenging to track failed tests.
To address this, I wrote a script to aggregate all failed tests into
one file and add additional information such as instruction name or
immediate value \cite[sh\_utils/diff\_logger.sh]{repoRV2BTOR}.

I have executed approximately 5,000,000 tests on this model without a
single failure, which leads me to conclude that my implementation is
correct.
